\section{Background \& Related Work}
\label{sec:background}

In this section,
we provide an overview of the legal framework
that must be tackled in data spaces,
as well as a description of ODRL and DPV
and how they have been used so far
to tackle these regulatory requirements.

\paragraph{Regulatory Compliance in Data Spaces}

Several branches of EU law must be considered
when implementing data spaces,
in particular when looking at the
enforcement of access and usage control.
One particular area of focus is
data-related legislation,
in particular the
General Data Protection Regulation (GDPR)~\cite{gdpr},
the Data Governance Act (DGA)~\cite{dga},
and the Data Act~\cite{data_act}, as
these regulatory instruments introduced by the European Commission
collectively shape the legal environment
for data sharing and governance in data spaces.
The GDPR establishes the foundational framework
for the protection of personal data,
defining principles such as lawfulness and purpose limitation,
and providing data subject rights
and obligations for data controllers,
which must be respected whenever personal data is processed,
including in cross-organisational data sharing scenarios typical of data spaces.
The DGA aims to facilitate trustworthy data sharing
by creating governance mechanisms that enable the reuse of
protected public sector data,
regulate data intermediation services,
and promote altruistic data sharing.
The Data Act complements this framework
by addressing access to and sharing of data
generated by connected products and services,
enabling broader data portability and sharing
across sectors to stimulate innovation and competition.
Together,
these regulations form a layered legal foundation
for European data spaces:
GDPR ensures protection of personal data,
while the DGA and the Data Act promote trusted mechanisms
and rights for data sharing and reuse,
enabling interoperable and legally compliant data ecosystems.
Additionally,
there is sectorial legislation
that should also be considered for the
implementation of data spaces.
For instance,
the European Health Data Space (EHDS)~\cite{ehds} regulation
establishes a common framework for the access,
exchange, and reuse of electronic health data,
forming the first sector-specific data space legislation
under the EU data strategy.
It aims to empower individuals with greater control and
cross-border access to their electronic health records,
while enabling the secondary use of health data for research,
innovation, and policy-making purposes.
Furthermore,
besides data-related and sector-based legislation,
other fields of law must be considered,
such as intellectual property rights or
competition law~\cite{data_spaces_support_centre_summary_2026}.

\paragraph{Access and Usage Policies Enforcement}

Besides being the \emph{de facto} standard
in data spaces for the expression of usage control policies,
ODRL was selected because it is a W3C standard specifically designed to represent machine-readable policies and rules governing the use of digital assets. 
At the core of the ODRL model~\cite{iannella_odrl_2018} is the Policy, which groups one or more Rules that define permissions, prohibitions, or obligations. Each rule specifies an Action performed by a Party over an Asset, and can be further refined using Constraints that restrict when or how the action may occur, expressed through left operands, operators, and right operands. Rules may also include Duties, representing additional actions that must be fulfilled when a rule is exercised. This model allows policies such as offers, agreements, or requests to be expressed in a machine-readable and interoperable manner.
However, since ODRL is domain-agnostic, DPV~\cite{pandit_data_2024} can be used alongside it to provide a controlled vocabulary of privacy and data protection concepts~---~such as legal bases, legal roles, and purposes for processing data~---~allowing policies to incorporate legally relevant semantics and regulatory concepts (e.g., from the GDPR).
In this context,
DPV is the \emph{de facto} standard for 
the representation of legal requirements
related to data protection,
as it is an open source solution
that provides the most complete set of taxonomies
for this domain~\cite{esteves_analysis_2024}.
DPV's core specification include taxonomies describing personal data categories, processing operations, and purposes for data use, as well as entities and roles such as data controllers, processors, and data subjects. Additional taxonomies describe legal concepts (e.g., laws and legal bases), technical and organisational measures, risks and risk mitigation measures, locations and jurisdictions, technologies and AI systems, and sector-specific contexts. Together, these vocabularies enable the interoperable representation of information about data processing activities and compliance requirements, which can then be incorporated into policy languages such as ODRL.

Previous ODRL profiles have looked at the combined usage of ODRL and
DPV~\cite{esteves_odrl_2021,pandit_enhancing_2024,golpayegani_aiup_2024},
however they did not provide a full overview and mapping of
all existing DPV taxonomies into ODRL constructs.
Furthermore, 
other works have used both
solutions~\cite{esteves_semantic_2025,esteves_fostering_2022,plazaortiz_authentication_2025,papagiannakopoulou_semantic_2025,yumusak_data_2024,de_vos_odrl_2019},
but failed to provide an explicit mapping of these technologies.
In this context,
this article fills this gap
by providing a complete DPV-ODRL mapping.

