\subsection{Limitations}
\label{sec:limitations}

While this work provides a structured and
comprehensive mapping between DPV and ODRL,
several limitations must be acknowledged.

First, the contribution primarily focuses on
conceptual alignment and specification artefacts,
including example policies and a SHACL-based validation template.
However,
it does not yet include a systematic implementation and evaluation
within realistic data space environments.
As such, the proposed profile has not been empirically assessed
in operational settings involving interoperable policy engines
or real-world data sharing scenarios.
Building on the latter,
the policy examples provided throughout the article
serve an illustrative purpose
but are not validated within concrete data space scenarios.
While they demonstrate how DPV concepts can be embedded within ODRL policies,
they do not account for challenges that may arise in real deployments,
such as policy conflicts, negotiation workflows,
or integration with existing enforcement mechanisms.
As a result,
the applicability of these examples to complex,
multi-party data sharing environments remains to be validated.
Finally,
it is important to consider that DPV is an actively evolving vocabulary
that is continuously extended to cover new data-related regulations,
sector-specific requirements, and emerging technologies.
While this extensibility is a strength,
it also introduces a moving target for the proposed mapping.
Hence,
changes or additions to DPV~--~such as
new legal concepts, jurisdictions, or domain-specific extensions~--~may
require ongoing updates to the DPV-ODRL alignment
to maintain completeness and consistency.